### 2019-02-27 paper/intro-notes.tex
Work on long-term effects of spacing
Combine intro

There is amble evidence that son preference has affected fertility decisions in India and 
elsewhere.
It is, for example, clear that a significant proportion of Indian families follow a
son preference based differential stopping behavior, where, for a given number of
children, they are more likely to have an additional child if they have not yet
reached their preferred number of sons 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.
A smilar pattern is found in many other countries 
\citep[see, for example][]{larsen98,filmer09,Altindag2016}.%
\footnote{
One potential effect of this behavior is that girls tend to end up in larger families, 
which may have negative impacts on survival and human capital investments, even in 
the absence of direct discrimination \citet{Jensen2003,Kugler2017}.
}

@@

In India, there is evidence that both the prior and the subsequent child have higher
likelihood of dying the closer they are spaced, although the effect is not symmetrical
\citep{Whitworth2002,Bhargava2003,Maitra2008,Makepeace2008}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explain a limited amount
of neonatal mortality.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, and El Salvadore that
short intervals between birth has negative health effects
\citep{Curtis1993,Davanzo2008,Gribble2009,Saha2013}.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again quicker, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.

[how is spacing achieved?]
This longer spacing 
\citep{Rahman1993}
\citep{Jayachandran2011}

@@

[spacing and outcomes - children]

[spacing and outcomes - mothers]

>>>>

There is amble evidence that son preference has affected fertility decisions in India and 
elsewhere.
A significant proportion of Indian families follow a son preference based differential 
stopping behavior, where, for a given number of children, they are more likely to have an 
additional child if they have not yet reached their preferred number of sons 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.
A smilar pattern is found in many other countries 
\citep[see, for example][]{larsen98,filmer09,Altindag2016}.%
\footnote{
One potential effect of this behavior is that girls tend to end up in larger families, 
which may have negative impacts on survival and human capital investments, even in 
the absence of direct discrimination \citet{Jensen2003,Kugler2017}.
}

@@

In India, there is evidence that both the prior and the subsequent child have higher
likelihood of dying the closer they are spaced, although the effect is not symmetrical
\citep{Whitworth2002,Bhargava2003,Maitra2008,Makepeace2008}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explain a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again quicker, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, and El Salvadore that
short intervals between birth has negative health effects
\citep{Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

The potential effects of spacing are not restricted to child outcomes, although
there is less evidence on longer term effects, and most of what we have comes from
Western countries.
In the US, for example, longer spacing between births increases the older sibling's
test scores on the Peabody Individual Achievement Test, although there is no
statistically significant effect on the younger sibling \citep{Buckles2012}.
In line with this, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.
These results are, however, not uniformly supported, with recent results from Sweden 
showing no relationship between birth spacing, even if very short, and outcomes such as 
years of education completed, earnings, and unemployment \citep{Barclay2017}.

The effect of spacing on mother's outcomes have received less attention than the
effect on children, and reviews of the literature show mixed results on mother's 
anthropometric status \citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are both
more likely to have short birth spacing and more likely to have anemia, possibly as
the result of the short spacing resulting in maternal depletion \citep{Milazzo2018}.
In Western countries, having a longer interval between births appear to have positive
effects on long-term labor market outcomes, such as participation and 
income \citep{Gough2017,Karimi2014}.

[how is spacing achieved?]
\citep{Rahman1993}
\citep{Jayachandran2011}
\citep{Alam2018}

@@

### 2019-02-27 paper/sex_selection_spacing-ver2-new-intro.tex
Combine intro

Parents' spacing of births has long served as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference is therefore often associated with shorter spacing after the 
births of girls than boys 
\citep{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997,Soest2012,Rossi2015}.
Shorter spacing is, in turn, associated with worse health outcomes for girls 
and mothers 
\citep{arnold98,Conde-Agudelo2000,Whitworth2002,Razzaque2005,Rutstein2005,Conde-Agudelo2006}.%
\footnote{
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{repetto72,ben-porath76b,Das1987,Arnold1997,arnold98,clark00,dreze01,filmer09,Basu2010,Altindag2016}.
}

However, the introduction of prenatal sex determination fundamentally changed the 
relationship between son preference and birth spacing.
Couples, who had shorter birth spacing before because of son preference,
now have access to prenatal sex determination and sex-selective abortions,
and each abortion increases birth spacing by six months to a year.
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.

As a result, we now have a situation where we may observer \emph{longer} spacing
for families with daughters than for families with sons, precisely because
strong son preference leads them to use sex selection.
Working in the opposite direction, couples with strong son preference
will likely still try to conceive earlier in the absence of sons. 
They may even shorten the time until conception, knowing that they might
have to go through multiple pregnancies and abortions before they
conceive a son.
To further complicate matters, we can still observe shorter birth spacing 
after the births of daughters as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

@@

At a broader level, we know less about what determines spacing behavior in 
developing countries than in developed countries.%
\footnote{
Even the effects of access to contraceptive on spacing are uncertain.
On the one hand, access to contraceptives allows women to avoid too 
short spacing between birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of
contraceptives can lead to shorter spacing between births 
\citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level 
of contraception, which results in longer spacing, to avoid having 
too many children by accident.
But, as contraception becomes more effective parents can more
easily avoid future births, allowing them to reduce the spacing 
between births without having to worry about overshooting their 
preferred number of children.
This idea may explain why better-educated women have
shorter spacing than less educated women in some instances
 \citep{Tulasidhar1993,Whitworth2002}.
These two counteracting effects may explain why finding statistically 
significant effects of contraception use on birth spacing is 
difficult \citep{Yeakey2009}.
}
With increasing numbers of women entering the labor force in developing
countries, understanding how couples make timing decisions will be necessary
for the design of suitable policies \citep{Portner2018}.
Furthermore, if spacing does affect health outcomes for mother and children, it is 
essential to understand what drives changes in spacing.
We know, for example, that health outcomes for girls appear to improve
in the presence of sex selection \citep{Lin2014,Hu2015}.
It is possible that these improvements are an unintended side-effect of
the longer spacing that arises from sex-selective abortions, rather than
because the smaller number of girls makes them more valued as is often
assumed.

In this paper, I introduce and apply a novel empirical method that directly incorporates 
the effects of sex-selective abortions on 
the duration between births
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
My proposed method allows for the time since the previous 
birth to affect the decision on sex selection.
By examining under what circumstances sex selection decisions change with 
spacing, we can draw a more nuanced picture of the degree of son preference.

% [Why India?]
I apply the method to birth histories of Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016. 
India is a particularly compelling case.
On the one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06}.%
\footnote{
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.
% %
% \footnote{
% This measure of son preference is commonly used in the literature. 
% See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
% }

There have also been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971 and still is.
The first reports of sex determination appeared around 
1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}
% Hence, the data make it possible to show how the spacing between births and
% sex ratios have changed with the introduction of prenatal sex determination,
% and whether the ban affected the relationship between birth spacing and sex 
% ratios.

>>>>

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
fundamentally change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion increases the interval between births.
As a result, we may now observe \emph{longer} average spacing for families with stronger 
son preference, precisely because it makes them more likely to use sex selection.
To further complicate matters, families with son preference who---for one reason or 
another---do not use prenatal sex selection may continue to have shorter birth spacing 
than those with a lower preference for sons.

In this paper, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce and apply an empirical method that directly incorporates the effects of 
sex-selective abortions on the duration between births 
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why India?]
India is a particularly compelling case.
India has long shown a strong male preference, especially in the northern states 
\citep{Kishor1993,murthi95,arnold98}.%
\footnote{
The proportion of couples who ideally wants more boys than girls does, however,
appear to be decreasing over time and with higher education \citep{bhat03,pande07}.
}
As a result, a significant proportion of Indian families follow a son preference based 
differential stopping behavior, where, for a given number of children, they are more 
likely to have an additional child if they have not yet reached their preferred number of 
sons \citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
A smilar pattern is found in many other countries 
\citep[see, for example][]{larsen98,filmer09,Altindag2016}.
}
Furthermore, higher mortality risk for females than males at most ages has lead to an 
almost continuous increase in India's overall ratio of males to females over the last 
century \citep{dyson01,Navaneetham2011,Bongaarts2015}.
Over the last three decades, India has also seen dramatic increases in the 
males-to-females ratio at birth as access to prenatal sex determination 
expanded \citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2012}.
In addition, fertility has declined substantially, to the point where it is now
at, or even below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
This decline in fertility, combined with the strong preference for sons, has lead to an 
intensified pressure to use sex selection \citep{Guilmoto2009a,Bongaarts2013,Jayachandran2017}.%
\footnote{
In some instances, however, the fertility decline appears to have created a stronger 
aversion to daughters rather than a stronger preference for son \citet{DiamondSmith2008}.
}

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, it has been widely used as a measure of son preference.
In India the sex of the last-born child significantly impact the duration to next birth,
with the expected birth interval about three percent longer if the last-born was son than
if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Similarly, ethic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are very small \citep{Pong1994}.
}
India is not alone in sex composition affecting subsequent birth spacing.
In both Bangladesh and Vietnam the more boys a family has the longer the expected duration 
to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia the evidence is more mixed.
North Africa shows shorter spacing in the absense of sons, while a similar pattern does
not exists in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

Second, birth spacing can have substantial impact on children's outcomes.
Most of the research finds a negative impact of short spacing on child health, although 
identifying the causal mechanisms have proven more difficult
\citep{Conde-Agudelo2006,Conde-Agudelo2012}.
Broadly speaking, there are three pathways that birth intervals can affect health.
First, maternal depletion, where the mother does not have sufficient
time to recover after a pregnancy.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiency.
}
Second, disease transmission both from mother to child and between children close in age.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is especially of interest when spacing decisions are driven by 
son preference.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to negatively affect the older child.

In India, there is evidence that both the prior and the subsequent child have higher
likelihood of dying the closer they are spaced, although the effect is not symmetrical
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Maitra2008,Makepeace2008}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explain a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again quicker, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, and El Salvadore that
short intervals between birth has negative health effects
\citep{Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

The potential effects of spacing are not restricted to early outcomes, although
there is less evidence on longer term effects, and most of what we have comes from
Western countries.
In the US, for example, longer spacing between births increases the older sibling's
test scores on the Peabody Individual Achievement Test, although there is no
statistically significant effect on the younger sibling \citep{Buckles2012}.
In line with this, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.
These results are, however, not uniformly supported, with recent results from Sweden 
showing no relationship between birth spacing, even if very short, and outcomes such as 
years of education completed, earnings, and unemployment \citep{Barclay2017}.

Third, the mother may also be affected by the length of the birth intervals.
The effect of spacing on mother's outcomes have received less attention than the
effect on children, and reviews of the literature show mixed results on mother's 
anthropometric status \citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are both
more likely to have short birth spacing and more likely to have anemia, possibly as
the result of the short spacing resulting in maternal depletion \citep{Milazzo2018}.
In Western countries, having a longer interval between births appear to have positive
effects on long-term labor market outcomes, such as participation and 
income \citep{Gough2017,Karimi2014}.

[Other reasons for studying spacing]
[Education and socio-economic status and spacing]
Finally, we know less about what determines spacing behavior in 
developing countries than in developed countries.
Even the effects of access to contraceptive on spacing are uncertain.
On the one hand, access to contraceptives allows women to avoid too 
short spacing between birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of
contraceptives can lead to shorter spacing between births 
\citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level 
of contraception, which results in longer spacing, to avoid having 
too many children by accident.
But, as contraception becomes more effective parents can more
easily avoid future births, allowing them to reduce the spacing 
between births without having to worry about overshooting their 
preferred number of children.
This idea may explain why better-educated women have
shorter spacing than less educated women in some instances
 \citep{Tulasidhar1993,Whitworth2002}.
These two counteracting effects may explain why finding statistically 
significant effects of contraception use on birth spacing is 
difficult \citep{Yeakey2009}.
With increasing numbers of women entering the labor force in developing
countries, understanding how couples make timing decisions will be necessary
for the design of suitable policies \citep{Portner2018}.

[how is spacing achieved?]
\citep{Rahman1993}
\citep{Jayachandran2011}
\citep{Alam2018}

Despite the potential important impact of birth spacing on outcomes for both
mother and children, there has so far been no examination of what effect access to
sex selection has had on spacing.
We know that health outcomes for girls appear to improve in the presence of sex 
selection \citep{Lin2014,Hu2015}.
It is possible that these improvements are an unintended side-effect of the longer spacing 
that arises from sex-selective abortions, rather than because the smaller number of girls 
makes them more valued as is often assumed.
A first step

[Predicted changes in spacing with sex selection]

[how much longer intervals]
 by six months to a year.
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.

Direct information on the use of sex selection is not available, but it is possible
to compare different periods as a proxy since there has been substantial changes in access 
and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971 and still is.
The first reports of sex determination appeared around 
1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}
[TK something on split 1995 to 2016 into two groups?]

[OLD VERSION FROM HERE]

@@
